Implementing simplified INLA into `aghq`

### 2022-11-22 src/dev_sinla/sinla.Rmd
plot_marginal_spline function, and old and new Laplace marginal methods

#' Build a Lagrange polynomial interpolant of the marginal posterior
ss <- splines::interpSpline(gg$nodes, lps, bSpline = TRUE, sparse = TRUE)

interpolant <- function(x) {
  as.numeric(stats::predict(ss, x)$y)
}

finegrid <- seq(-5, 5, by = 0.1)

df <- data.frame(x = finegrid, y = exp(interpolant(finegrid)))

ggplot(df, aes(x = x, y = y)) +
  geom_line() +
  theme_minimal() +
  labs(x = "phi_prev[1]", y = "Posterior")
```

@@

Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
We want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z),
$$
so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.

@@

As before, the weight $\omega(z)$ is:

```{r}
modesandhessians$weights[z]
```

And also as before, the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:

```{r}
#' All of the theta
modesandhessians[, thetanames]

#' Just the zth one
theta <- as.numeric(modesandhessians[z, thetanames])
theta

- h_norm(theta)
```

Now, getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:

@@

#' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())

@@

obj_fixed_theta_norm(x_test_value)
```

Putting it together:

```{r}
#' On the log-scale
as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])

#' Unlogged
exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
```

Sum this over `z`:

```{r}
pdf <- 0

@@

  #' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
  quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())

  pdf <- pdf + exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
}

pdf
```

@@

The method above for doing quadrature over the full Laplace approximation should work, but it requires a lot of computation.
Instead, we might be able to do this more cost effectively by integrating $\theta$ out before normalising $\tilde p_\text{LA}(x_i, \theta(z), y)$.
This would look like first obtaining
$$
\tilde p(x_i, y) = \sum_z \tilde p_\text{LA}(x_i, \theta(z), y) \omega(z),
$$
then normalising by the previously obtained $\tilde p_\text{AQ}(y)$ as follows
$$
\tilde p(x_i \, | \, y) = \frac{\tilde p(x_i, y)}{\tilde p_\text{AQ}(y)}.
$$

```{r}
#' Always (?) want the same initialisation parameters (though they probably could be better right?)
param_i_fixed_theta <- list(
  beta_prev = 0, 
  phi_prev_i = 0, 
  phi_prev_minus_i = rep(0, data$n - 1),
  log_sigma_phi_prev = 0
)

#' Always fixing the hyperparameters
map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))

#' Going to store the functions p_LA(x_i, theta(z), y) in objs
objs <- list()

@@

  objs[[z]] <- obj_fixed_theta
}

aghq_laplace_marginal <- function(x) {
  pdf <- 0
  for(z in 1:nrow(modesandhessians)) {
    pdf <- pdf + exp(as.numeric(-objs[[z]]$fn(x))) * modesandhessians[z, "weights"]
  }
  return(pdf)
}

#' Issue is that we do the normalisation after now..
aghq_laplace_marginal(x = x_test_value)
aghq_laplace_marginal(x = -1)
```

>>>>

plot_marginal_spline <- function(lps) {
  #' Build a Lagrange polynomial interpolant of the marginal posterior
  ss <- splines::interpSpline(gg$nodes, lps, bSpline = TRUE, sparse = TRUE)
  interpolant <- function(x) { as.numeric(stats::predict(ss, x)$y) }
  finegrid <- seq(-5, 5, by = 0.1)
  df <- data.frame(x = finegrid, y = exp(interpolant(finegrid)))

  ggplot(df, aes(x = x, y = y)) +
    geom_line() +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")
}

plot_marginal_spline(lps)
```

@@

As before, but with the full Laplace approximation in place of the Gaussian.
Much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$ in
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
$$

@@

As before, the weight $\omega(z)$ is `modesandhessians$weights[z]` and the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is `- h_norm(as.numeric(modesandhessians[z, thetanames]))`.
Getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:

@@

#' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
quad <- aghq::aghq(ff = obj_fixed_theta, k = 7, startingvalue = 0, control = aghq::default_control_tmb())

@@

- obj_fixed_theta_norm(x_test_value)
```

Putting it together on the log-scale:

```{r}
as.numeric(- obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
```

Sum over `z`:

```{r}
old_laplace_marginal <- function(x) {
  lp <- vector(mode = "numeric", length = nrow(modesandhessians))

@@

    quad <- aghq::aghq(ff = obj_fixed_theta, k = 7, startingvalue = 0, control = aghq::default_control_tmb())
    obj_fixed_theta_norm <- function(xx) obj_fixed_theta$fn(xx) + quad$normalized_posterior$lognormconst

    lp[z] <- as.numeric(- obj_fixed_theta_norm(x)) + (- h_norm(theta)) + log(modesandhessians$weights[z])
  }

old_laplace_marginal(x_test_value)

lps <- sapply(gg$nodes, old_laplace_marginal)
plot_marginal_spline(lps)
```

@@

The method above should work, but it requires a lot of computation.
Instead, we might be able to do this more cost effectively by integrating $\theta$ out *before* normalising $\tilde p_\text{LA}(x_i, \theta(z), y)$.
This would look like first obtaining
$$
\tilde p(x_i, y) = \sum_z \tilde p_\text{LA}(x_i, \theta(z), y) \omega(z),
$$
then normalising by the previously obtained $\tilde p_\text{AQ}(y)$ as follows
$$
\tilde p(x_i \, | \, y) = \frac{\tilde p(x_i, y)}{\tilde p_\text{AQ}(y)}.
$$

```{r}
#' Make sure the log normalising constant is available
quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
quad$normalized_posterior$lognormconst

laplace_marginal <- function(x) {
  lp <- vector(mode = "numeric", length = nrow(modesandhessians))

@@

    lp[z] <- as.numeric(- obj_fixed_theta$fn(x)) + log(modesandhessians$weights[z])
  }

  matrixStats::logSumExp(lp) - quad$normalized_posterior$lognormconst
}

laplace_marginal(x_test_value)

lps <- sapply(gg$nodes, laplace_marginal)
plot_marginal_spline(lps)
```
